---
title: "methylclock quality-control report"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
params:
  qc: null
  plots: null
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center")
qc <- params$qc
plots <- params$plots
has_age <- isTRUE(qc$has_age)
note <- function(section) {
    txt <- qc$notes[[section]]
    if (!is.null(txt)) cat(paste(txt, collapse = " "), "\n\n")
}
# Grow the figures with the number of clocks so the panels keep their shape
# instead of being squeezed into a fixed height.
n_shown <- max(1L, length(qc$shown))
h_dist <- min(22, max(4, ceiling(n_shown / 4) * 1.8 + 1))
h_corr <- min(14, max(4.5, n_shown * 0.32 + 2.5))
```

Every section below states what it shows, what it reads as on your data, and
what produces such a pattern. It describes; the interpretation is yours to
make. Terms in the tables are defined in the glossary at the end.

## Summary

```{r summary, results = "asis"}
cat(paste0("- ", qc$commentary, "\n"), sep = "")
```

## Samples

```{r samples}
tab <- data.frame(Samples = qc$samples$n)
if (has_age) {
    a <- qc$samples$age
    tab <- cbind(tab, `Min age` = round(a[["Min."]], 1),
                 `Median age` = round(a[["Median"]], 1),
                 `Max age` = round(a[["Max."]], 1))
}
knitr::kable(tab, row.names = FALSE, check.names = FALSE)
if (!is.null(qc$groups))
    knitr::kable(as.data.frame(qc$groups), col.names = c("Group", "N"),
                 row.names = FALSE)
```

```{r samples-note, results = "asis"}
note("variables")
```

```{r coverage-head, eval = !is.null(qc$coverage), results = "asis"}
cat("## CpG coverage\n\n")
note("coverage")
```

```{r coverage-table, eval = !is.null(qc$coverage)}
knitr::kable(qc$coverage, row.names = FALSE,
             col.names = c("Clock", "CpGs in clock", "Present", "% present"))
```

```{r missing-head, eval = !is.null(qc$missing_by_clock), results = "asis"}
cat("## Missing values filled in per clock\n\n")
note("missing")
```

```{r missing-table, eval = !is.null(qc$missing_by_clock)}
knitr::kable(qc$missing_by_clock, row.names = FALSE,
             col.names = c("Clock", "CpGs present", "Missing values",
                           "% missing"))
```

```{r agree-head, eval = !is.null(plots$correlation), results = "asis"}
cat("## Agreement between clocks\n\n")
note("agreement")
```

```{r plot-corr, eval = !is.null(plots$correlation), fig.width = 7, fig.height = h_corr}
plots$correlation
```

```{r dist-head, eval = !is.null(plots$distributions), results = "asis"}
cat("## Distribution of the estimates\n\n")
note("distributions")
```

```{r plot-dist, eval = !is.null(plots$distributions), fig.width = 8, fig.height = h_dist}
plots$distributions
```

```{r acc-head, eval = has_age, results = "asis"}
cat("## Accuracy against chronological age\n\n")
note("accuracy")
```

```{r plot-age, eval = !is.null(plots$age), fig.width = 8, fig.height = 5.5}
plots$age
```

```{r accuracy, eval = has_age && !is.null(qc$accuracy)}
knitr::kable(qc$accuracy, row.names = FALSE)
```

```{r ba-head, eval = !is.null(plots$bland_altman), results = "asis"}
cat("### Agreement with age (Bland-Altman)\n\n")
note("bland_altman")
```

```{r plot-ba, eval = !is.null(plots$bland_altman), fig.width = 8, fig.height = 5.5}
plots$bland_altman
```

```{r sample-head, eval = !is.null(qc$sample_qc), results = "asis"}
cat("## Per-sample quality\n\n")
note("samples")
```

```{r sample-table, eval = !is.null(qc$sample_qc)}
sq <- qc$sample_qc
flagged <- sq[sq$is_outlier %in% TRUE, , drop = FALSE]
flagged <- flagged[order(-flagged$outlier_score), , drop = FALSE]
if (nrow(flagged))
    knitr::kable(utils::head(flagged, 20), row.names = FALSE)
```

```{r sample-none, eval = !is.null(qc$sample_qc), results = "asis"}
if (!any(qc$sample_qc$is_outlier %in% TRUE))
    cat("No samples were flagged.\n")
```

```{r no-age, eval = !has_age, results = "asis"}
cat("## Accuracy\n\nNo chronological age was supplied, so the error against",
    "age (`R2`, `MAE`), the bias and the age acceleration are not part of this",
    "report. The clocks were still computed, and the coverage, agreement and",
    "distribution sections above do not need ages.\n")
```

## Glossary

| Term | What it means |
|---|---|
| **CpG coverage** | The share of the CpG sites a clock uses that are present in your matrix. |
| **Imputed (filled-in) value** | A value that was missing in the matrix and was replaced before the clock ran, using the method named above. It is a stand-in, not a measurement. |
| **r** | Pearson correlation between a clock's estimate and chronological age: how tightly the estimate follows age, regardless of scale. Runs from -1 to 1. |
| **R2** | The square of *r*: the share of the variation in age that the estimate accounts for. |
| **MAE** | Median absolute error: the typical size of the gap between estimate and chronological age, in years. Unlike *r*, it is affected by a constant offset. |
| **Bias** | The average signed gap between estimate and age. Positive means the clock reads older than the calendar age. |
| **Limits of agreement** | The band holding 95% of the individual gaps between estimate and age (the dashed lines in the Bland-Altman plot). |
| **z-score** | How many standard deviations a value sits from the mean of the dataset. Used here to describe how far a sample is from the rest. |
| **Age acceleration** | The part of a clock's estimate not explained by chronological age: the residual of the estimate regressed on age. |

<hr>
<small>Generated by the methylclock package.</small>
